Supermartingales and Termination Analysis of Statistical PCF

### 2020-09-29 probabilisticProgrammingMartingales.tex
Continue proof of canonicalCousins lemma.

\begin{proof}
A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underscore f_i$ and $\underscore f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order. Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur. It is only required that the reductions that do occur occur in the correct order.
\lo{This paragraph (especially the last sentence) is ambiguous.
What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
\iffalse
\lo{So this is permitted?
\[
(II)(I (I \bar I)) \to
(II)(I\bar I) \to
(II)\bar I 
\]
but not this: 
\(
(II)(I (I \bar I)) \to
(II)(I \bar I) \to
(II)\bar I 
\)?}
\fi

@@

The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possible overlapping. If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order. This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$. As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.

In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$. There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$. Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps. Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the position of the first relevant reduction in $Y$ be $\delta$. Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ results in $\gamma, \beta_k, \dots, \beta_1$. In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$.

@@

    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2, $\alpha < \beta_1$ and $\gamma$ is right of $\alpha$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by one of the cases b to f in the definition of $\sim_c$, then swapping $\gamma$ down past $\alpha$ by case a, then down past all the images of $\beta_1$, also by case a. Eventually it reaches some $X_i$ where the relevant sub-reduction-sequence is $\alpha$ then whatever $\beta_1$ was turned into in $X \sim_c X_1$ then $\gamma$, before moving on to swapping pairs earlier in the reduction sequence. The sequence $(Y_j)_j$ starts by swapping $\gamma$ down past $\alpha$ by case a, then swapping $\alpha$ and $\beta_1$ as above. The subskeleton at $\alpha$ at the relevant point in the reduction sequence if the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.

@@

>>>>

\begin{proof}
A reduction sequence is in call-by-value order if for every pair of reductions in the sequence at positions $\alpha$ and $\beta$, where the reduction at $\alpha$ happens first, either $\alpha \leq \beta$ or $\alpha$ and $\beta$ are disjoint, with $\alpha$ to the left of $\beta$ (i.e.~the first elements of the sequences $\alpha, \beta$ where they differ are $@_1$ and $@_2$, $\underline f_i$ and $\underline f_j$ for $i < j$, or $\textsf{if}_i$ and $\textsf{if}_j$ for $i < j$, in that order. Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur. It is only required that the reductions that do occur occur in the correct order.
\lo{This paragraph (especially the last sentence) is ambiguous.
What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
\iffalse
\lo{So this is permitted?
\[
(II)(I (I \bar I)) \to
(II)(I\bar I) \to
(II)\bar I 
\]
but not this: 
\(
(II)(I (I \bar I)) \to
(II)(I \bar I) \to
(II)\bar I 
\)?}
\fi

@@

The remaining cases are when the subsequence of reduction steps involved in $Y \sim_c X$ occur before the last out-of-order pair in $X$, but possibly overlapping. If there is no overlap, then the $\sim_c$ steps do not interfere with each other at all, and can simply be performed in the other order. This gives a different sequence of $\sim_c$ steps $Y \sim_c Y_1 \sim_c X_1 \sim_c^* X_{cbv}$. As $Y_{cbv} = (Y_1)_{cbv}$, we can proceed by induction in this case and use the fact that $(Y_1)_{cbv} = (X_1)_{cbv} = X_{cbv}$ to acheive the result.

In the other case, there is some overlap between the regions involved in $Y \sim_c X$ and $X \sim_c X_1$, but the last reduction in $X \sim_c X_1$ is later than the last reduction in $Y \sim_c X$. There are only 2 reductions in $X$ involved in $X \sim_c X_1$, therefore the region involved in $Y \sim_c X$ ends on the first of the reductions involved in $X \sim_c X_1$. Now consider the specific sequence of positions of the reductions in $X$ involved in either of these steps. Let the last of these positions be $\alpha$, the first $\gamma$, and the second-last, third-last and so on $\beta_1, \beta_2, \dots$ respectively, and let the position of the first relevant reduction in $Y$ be $\delta$. Then $X \sim_c X_1$ swaps $\beta_1$ and $\alpha$, and $\alpha$ comes before $\beta_1$ in CBV order (i.e.~$\alpha < \beta_1$ or $\alpha$ is left of $\beta_1$), and the step $Y \sim_c X$ either swaps $\delta$ and $\gamma$ resulting in $\gamma, \beta_k, \dots, \beta_1$, or swaps $\delta$ with the next (or several next) reduction (s) resulting in $\gamma, \beta_1$, depending on the direction of the $\sim_c$ step. In each of the cases below, it is required to show that the canonical sequence $(X_i)_i$ eventually reaches some term which is the same as one in $(Y_j)_j$, and from that point on, they match, therefore they reach the same end result: $X_{cbv} = Y_{cbv}$. Not all of the cases below are disjoint, but they do cover all of the remaining possibilities.

@@

    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2, $\alpha < \beta_1$ and $\gamma$ is right of $\alpha$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by one of the cases b to f in the definition of $\sim_c$, then swapping $\gamma$ down past $\alpha$ by case a, then down past all the images of $\beta_1$, also by case a. Eventually it reaches some $X_i$ where the relevant sub-reduction-sequence is $\alpha$ then whatever $\beta_1$ was turned into in $X \sim_c X_1$ then $\gamma$, before moving on to swapping pairs earlier in the reduction sequence. The sequence $(Y_j)_j$ starts by swapping $\gamma$ down past $\alpha$ by case a, then swapping $\alpha$ and $\beta_1$ as above. The subskeleton at $\alpha$ at the relevant point in the reduction sequence is the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.

@@

    \item $\alpha$ is disjoint from $\gamma$ and all the $\beta_i$: In this case, $\alpha$ is left of all the $\beta_i$, and if it isn't also left of $\gamma$, then $Y_1 = X$ and we're done. Otherwise, $(X_i)_i$ proceeds by swapping each of the $\beta_i$ in turn then $\gamma$ below $\alpha$ by case a, and $(Y_j)_j$ swaps everything up to $\delta$ to below $\alpha$ by case a, then performs the same swap as in $Y \sim_c X$, but after $\alpha$ in the reduction sequence, yielding $\alpha, \gamma, \beta_k, \dots, \beta_1$, which matches $X_{k+1}$.
    \item $\alpha > \gamma$: In this case, $Y_1 = X$.
